# Perception and Misperception of Clustering in Nonlinear Dimension Reduction: A User Study {#sec-second-paper}

Non-linear dimension reduction (NLDR) methods such as tSNE, UMAP, PHATE, TriMAP, and PaCMAP generate low-dimensional representations of high-dimensional data. However, identifying meaningful low-dimensional representations can be challenging due to factors such as closely spaced clusters. In this study, we use data simulated with the `cardinalR` package containing three clusters of different shapes to assess how well NLDR methods preserve inter-cluster structure. We focus on a specific metric; Between-to-within (BW) ratio as an indicator of global separation. The results reveal that UMAP and tSNE most effectively maintain meaningful global structure by preserving larger inter-cluster distances. PHATE, in contrast, emphasizes local connectivity, sometimes at the expense of global separation. PaCMAP and TriMAP appear less responsive to inter-cluster distance in this context.

```{r include=FALSE}
# Set up chunk for for knitr
knitr::opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
```

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tidyverse)
library(lme4)
library(broom.mixed)

library(ggplot2)
library(ggbeeswarm)
library(readr)
library(tidyr)
library(emmeans)

library(patchwork)
library(colorspace)
library(kableExtra)

library(conflicted)
library(binom)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
```

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     #aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "bottom", 
     legend.title = element_blank(), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm')
   )
)
```


